---
title: "Dusendang Graticule Biostats Sample"
author: "Jenny Dusendang"
date: "September 22, 2023"
output: html_document
---

## Background (from readme)

Consider a cohort study design with a defined exposure and an outcome of interest (which is not 
mortality).

- Study period from July 1st though Dec 31st 2022
- Exposure enters a subject into the exposed cohort and serves as the index date
- Exposed subjects are exact matched with unexposed subjects based on age, gender, and a healthcare 
encounter within the same week
- Matching is performed without replacement
- Exposed subjects may serve as unexposed matches during any week before the week of their exposure 
date

The provided tables (in csv format) contain the following data elements:
Patient table

- Patient ID
- Gender
- Date of birth
- Date of death | NULL

Encounter table
- Patient ID
- Encounter date
- Exposure binary indicator
- Outcome binary indicator

For each of the following analysis steps, provide working code (in a language of your choosing):

1. Load the data tables into dataframe(s) or a database
2. Compute median age at index date for the exposed cohort (including CI at 95% confidence level)
3. Compute aggregate time at risk for the exposed cohort, in person-days
4. Finally, provide pseudo-code to select a 1:1 unexposed cohort based on the method outlined above.


## Load packages
```{r}

library(dplyr)
library(misty)

```


## Load data 
```{r}

patient_raw = read.csv('patient_table.csv')
encounter_raw = read.csv('encounter_table.csv')

```


## Prep and clean 
```{r}

#are columns the right class? 
sapply(patient_raw, class)
sapply(encounter_raw, class)

#dates need to be date class and binaries need to be TRUE FALSE
patient = patient_raw %>%
  mutate(birth_date = as.Date(birthdate, format = '%Y-%m-%d'),
         death_date = as.Date(deceaseddate, format = '%Y-%m-%d'))

encounter = encounter_raw %>%
  mutate(encounter_date = as.Date(encounterdate, format = '%Y-%m-%d'),
         exposure2 = as.logical(exposure),
         outcome2 = as.logical(outcome))

```


## Prep exposed data 
```{r}

cohort = patient %>%
  #first outcome date
  left_join(encounter %>%
              filter(outcome2 == TRUE) %>%
              select(patientid, outcome_date = encounter_date) %>%
              group_by(patientid) %>%
              filter(outcome_date == min(outcome_date)) %>%
              ungroup(), by = 'patientid') %>%
  #first exposure date
  left_join(encounter %>%
              filter(exposure2 == TRUE) %>%
              select(patientid, exposure_date = encounter_date) %>%
              group_by(patientid) %>%
              filter(exposure_date == min(exposure_date)) %>%
              ungroup(), by = 'patientid')

exposed = cohort %>%
  #filter to exposed 
  filter(!is.na(exposure_date)) %>%
  #remove those with outcome prior to exposure (not 'at risk' for outcome)
  filter(is.na(outcome_date) | outcome_date > exposure_date)

```


## Median age at index for exposed (with 95% CI)
```{r}

age_index = exposed %>%
  mutate(age = as.numeric(exposure_date - birth_date) / 365.25)

#median 
median = median(age_index$age)
ci_median = ci.median(age_index$age, alternative = 'two.sided', conf.level = 0.95)

#print it 
print(paste('Median age at index among exposed: ', round(median, 1), ' years (95% CI: ', 
            round(ci_median$result$low, 1), '-', round(ci_median$result$upp, 1), ')', sep = ''))

```


## Time at risk for exposed 
Includes all time exposed before outcome, death, or end of study 
```{r}

time_at_risk = exposed %>%
  mutate(end_of_fu = pmin(outcome_date, death_date, as.Date('2022-12-31', format = '%Y-%m-%d'), 
                          na.rm = TRUE),
         time_at_risk = as.numeric(end_of_fu - exposure_date)) 

tot_time_at_risk = sum(time_at_risk$time_at_risk)

print(paste('Total time at risk among exposed: ', tot_time_at_risk, ' person-days', sep = ''))

```


## Matching 
```{r}

#create clean dataset of exposed patients
exposed_for_matching = exposed %>%
  select(exposed_patientid = patientid, exposed_index_date = exposure_date, 
         exposed_birth_date = birth_date, exposed_gender = gender)

#create clean dataset of unexposed patients 
unexposed = cohort %>%
  #when are unexposed no longer valid as matching potential 
  mutate(unexposed_end_date = pmin(exposure_date, death_date, na.rm = TRUE)) %>%
  select(unexposed_patientid = patientid, unexposed_end_date, unexposed_birth_date = birth_date,
         unexposed_gender = gender)

#grab all encounters 
unexposed_encounters = encounter %>%
  select(patientid, unexposed_encounter_date = encounter_date) %>%
  distinct()

#merge unexposed to encounters 
unexposed_options = unexposed %>%
  left_join(unexposed_encounters, by = c('unexposed_patientid' = 'patientid'))

#empty dataframe for matches to sit in
matches = data.frame()

#randomize order of datasets for matching (first arrange by ID then randomize) 
set.seed(5678)
exposed_for_matching = sample_n(exposed_for_matching %>% 
                                  arrange(exposed_patientid), 
                                nrow(exposed_for_matching))

set.seed(5678)
unexposed_options = sample_n(unexposed_options %>% 
                               arrange(unexposed_patientid, unexposed_encounter_date), 
                             nrow(unexposed_options))

#run the matching!
for (exposed_pats in 1:nrow(exposed_for_matching)){
  
  exposed_pat_row = exposed_for_matching %>%
    filter(row_number() == exposed_pats)
  
  exposed_patientid = pull(exposed_pat_row, exposed_patientid)
  exposed_gender = pull(exposed_pat_row, exposed_gender)
  exposed_birth_date = pull(exposed_pat_row, exposed_birth_date)
  exposed_index_date = pull(exposed_pat_row, exposed_index_date)
  
  test_unexposed = unexposed_options %>%
    filter(
      #not the same person 
      unexposed_patientid != exposed_patientid &
        #same gender
        unexposed_gender == exposed_gender & 
        #match within 10 years of age
        as.numeric(abs(unexposed_birth_date - exposed_birth_date)) < 10 * 365.25 &
        #match must still be unexposed and alive at index
        unexposed_end_date > exposed_index_date &
        #must have an encounter in week surrounding index
        as.numeric(exposed_index_date - unexposed_encounter_date) < 7 &
        as.numeric(exposed_index_date - unexposed_encounter_date) >= 0) 
  
  #if there isn't a match move on 
  if (nrow(test_unexposed) == 0){
    
    matches = matches %>%
      bind_rows(exposed_pat_row %>%
                  select(exposed_patientid))
  
    #if there is a match put it in the matches dataset   
  }else{
    
    picked_match = test_unexposed %>%
      filter(row_number() == 1) %>%
      select(unexposed_patientid)
    
    matches = matches %>%
      bind_rows(exposed_pat_row %>% 
                  select(exposed_patientid) %>%
                  bind_cols(picked_match))
    
    #remove patient from the pool
    unexposed_options = unexposed_options %>%
      filter(unexposed_patientid != pull(picked_match, unexposed_patientid))
    
  }
  
}

#probably need to reconsider matching criteria as we didn't get many matches 

```




