---
title: "Dusendang Graticule Biostats Sample"
author: "Jenny Dusendang"
date: "September 22, 2023"
output: html_document
---

## Background (from readme)

Consider a cohort study design with a defined exposure and an outcome of interest (which is not 
mortality).

- Study period from July 1st though Dec 31st 2022
- Exposure enters a subject into the exposed cohort and serves as the index date
- Exposed subjects are exact matched with unexposed subjects based on age, gender, and a healthcare 
encounter within the same week
- Matching is performed without replacement
- Exposed subjects may serve as unexposed matches during any week before the week of their exposure 
date

The provided tables (in csv format) contain the following data elements:
Patient table

- Patient ID
- Gender
- Date of birth
- Date of death | NULL

Encounter table
- Patient ID
- Encounter date
- Exposure binary indicator
- Outcome binary indicator

For each of the following analysis steps, provide working code (in a language of your choosing):

1. Load the data tables into dataframe(s) or a database
2. Compute median age at index date for the exposed cohort (including CI at 95% confidence level)
3. Compute aggregate time at risk for the exposed cohort, in person-days
4. Finally, provide pseudo-code to select a 1:1 unexposed cohort based on the method outlined above.




## Load packages
```{r}

library(dplyr)
library(misty)

```




## Load data 
```{r}

patient_raw = read.csv('patient_table.csv')
encounter_raw = read.csv('encounter_table.csv')

```




## Prep and clean 
```{r}

#are columns the right class? 
sapply(patient_raw, class)
sapply(encounter_raw, class)

#dates need to be date class and binaries need to be TRUE FALSE
patient = patient_raw %>%
  mutate(birth_date = as.Date(birthdate, format = '%Y-%m-%d'),
         death_date = as.Date(deceaseddate, format = '%Y-%m-%d'))

encounter = encounter_raw %>%
  mutate(encounter_date = as.Date(encounterdate, format = '%Y-%m-%d'),
         exposure2 = as.logical(exposure),
         outcome2 = as.logical(outcome))

```




## Median age at index for exposed (with 95% CI)
Seems a little odd to calculate 95% interval for median so I'll do it for median and mean just to be 
safe 
Exposure date is index date 
```{r}

#do patients show up as exposed more than once? 
index_test = encounter %>% 
  filter(exposure2 == TRUE) %>%
  group_by(patientid) %>%
  summarize(count = n())

#yes, use first instance as the start of exposure 
index = encounter %>%
  filter(exposure2 == TRUE) %>%
  group_by(patientid) %>%
  filter(encounter_date == min(encounter_date)) 

#used this to check if pats had more than one first exposure visit (they did not) 
# %>%
#   summarize(count = n())

age_index = index %>%
  left_join(patient, by = 'patientid') %>%
  mutate(age = as.numeric(encounter_date - birth_date) / 365.25)

#mre standard mean calculation
mean = mean(age_index$age)
ci_mean = t.test(age_index$age)$conf.int

#median 
median = median(age_index$age)
ci_median = ci.median(age_index$age, alternative = 'two.sided', conf.level = 0.95)

#print it 
print(paste('Median age at index: ', round(median, 1), ' years (95% CI: ', 
            round(ci_median$result$low, 1), '-', round(ci_median$result$upp, 1), ')', sep = ''))

```



